---
title: "Tess McCabe's Lab 1"
output:
  pdf_document: default
  html_notebook: default
  html_document: default
---

```{r, echo=FALSE}
library(ggplot2)
library(tidyverse)

```




*1.	Evaluate the following:*
   a.	ln(1)
```{r, echo=TRUE}
log(1)
```
   b.	ln(0)
```{r, echo=TRUE}
log(0)
```
   c.	ln(e)
```{r, echo=TRUE}
log(exp(1)) 
```
   d.	ln(-5)
```{r, echo=TRUE}
log(5*(-1)) # Can't do that. 
```
   e.	-ln(5)
```{r, echo=TRUE}
-log(5)
```
   f.	ln(1/5)
```{r, echo=TRUE}
log(1/5)
```



g.	How does R represent when the output of a function is not a number?

R stores things in objects. If the output of a function isn't a number, it could be text output ie: 

```{r, echo=TRUE}
print("Hello world")
```

The output could be an object that stores numbers in an ordered way like a vector, matrix, dataframe or array ie: 
```{r, echo= TRUE}
vector<-colSums(iris[,1:3], na.rm = TRUE,dims = 1)
vector
```

The output could also be a comniation of any of those things in the form of a list, or a custom object class. 


*2.	Why are we multiplying the results of the atan and atan2 functions by 180/pi?*

Becuase R default functions take radians ot degrees, and we wanted our answers in degrees. 

*3.	How is the atan2 function different from the atan function?* 

The atan2() function retunrs the angle between the x axis an the line that conencts the origin to the  (x,y) coordinates. It's the same as atan(y/x). 

*4.	What is the difference between log and log10?*

log() the R funnction is what most people think of as natural log  aka $ln()$. log10 is log base ten, or what most people think of when they see $log(x)$. 

*5.	Given a right triangle with sides x = 5 and y = 13, calculate the length of the hypotenuse (show code)*


$$x^2+y^2=c^2$$

```{r, echo=TRUE}
x<-5
y<-13
c_squared<-5^2+13^2
c<-c_squared^(1/2)
c
```

*6.	If a population starts with a density of 5 individuals/hectare and grows exponentially at a growth rate r=0.04/year, what is the size of the population in π years? (show code)*

```{r, echo=TRUE}
density<-5 #indiv/hect
rate<- 0.04 # year^-1
years<-pi

population<-density+(rate*years)
population
```

*7.	Subtract the month you were born from the remainder left after you divide the year you were born by the day you were born (show code)*

```{r, echo=TRUE}
day_born<-23 #Fyi, tomorrow is my birthday. 
month_born<-1
year_born<-1995 
remainder<-1995%%23
remainder
month_born-remainder

```





*8.	Describe the difference in output between sum and cumsum.*

```{r, echo=TRUE}
sum(1:10) 
cumsum(1:10)
```
sum() returns a scalar, or a single number that represents the sum of all the values in a vector. cumsum() returns anouther vector that is the cumulative sum of all previous numbers up untill that point. 

*9.	Generate a sequence of even numbers from -6 to 6*
```{r, echo=TRUE}
x<--6:6
x
```
*10.	Generate a sequence of values from -4.8 to -3.43 that is length 8 (show code)*
```{r, echo=TRUE}
sequence<-seq(-4.8,-3.43,length=8)
sequence
```
a.	What is the difference between values in this sequence?

```{r, echo=TRUE}
diff(sequence)
```

b.	What is the sum of the exponential of this sequence?
```{r, echo=TRUE}
sum(exp(sequence)) #is this what you mean by the exponential? 
```
*11.	Calculate a second difference [a difference of differences] for the sequence 1:10 (show code)*

```{r, echo=TRUE}
sq<-1:10
dif<-diff(sq)
diff(dif) #zero, as one would expect
```

```{r, echo=TRUE}
dat = read.table("data/Lab1_frogs.txt",header=TRUE)
write.table(dat,"my_frogs.csv",row.names=FALSE,sep=",")
save(dat,x,y,z,file="Lab1.RData")
load("Lab1.RData")
#save.image("Lab1_all.RData") #I'm not going to run this one

write.table(dat, "Lab1_table.txt", row.names = FALSE, sep="|")

```

Text from table: 
"frogs"|"tadpoles"|"color"|"spots"
1.1|2.03698175474231|"red"|TRUE
1.3|2.87623092770957|"red"|FALSE

```{r, echo=TRUE}
dat2<-read.table("Lab1_table", sep = "|", header = TRUE)
dat2
```


*13.	Show just the spots data as characters*

```{r, echo=TRUE}
as.character(dat[,4])

```
*14.	Show the 3rd through 8th rows of the 1st  though 3rd columns*
```{r, echo=TRUE}
dat[3:8,1:3]

```
*15.	Show the first 3 rows*
```{r, echo=TRUE}
dat[1:3,]
```



*16.	Create a character vector that contains the names of 4 super heros.*

```{r, echo=TRUE}
x<-c("Ruth Bader Ginsburg", "Sonia Sotomayor", "Andrea Gibson", "Anne Bronte")
x
```

*17.	Show just the odd numbered rows in the frog data. Write this code for the GENERAL CASE (i.e. don’t just type c(1,3,5,…) but use functions that you learned in previous sections to set up the sequence.*



*18.	For the frog data set:* 
  a.	display just the rows where frogs have spots
  
```{r, echo=TRUE}

dat[dat$spots==TRUE,]
```
  b.	display just the rows where frogs are blue
```{r, echo=TRUE}
dat[dat$color=="blue",]

```
  c.	how many blue tadpoles are there?
```{r, echo=TRUE}
sum(dat[dat$color == "blue",]$tadpoles)
```
  d.	create a new object containing just the rows where there are between 3 and 5 tadpoles
```{r, echo=TRUE}
few_tadpoles<-dat[dat$tadpoles > 3 & dat$tadpoles < 5,]
few_tadpoles
```
  e.	display just the rows where there are less than 2.5 red frogs

```{r, echo=TRUE}
dat[dat$frogs <2.5 & dat$color=="red",]
```
  f.	display where either frogs do not have spots or there are more than 5 frogs


```{r, echo=TRUE}
dat[dat$frogs >5 | dat$spots=="FALSE",]
```



*19. Plot the following lines from 0 to 3 (hint: define x as a sequence with a small step size). Make sure to make the resolution of x sufficiently small to see the curves* 
  a.	ln(x) 
  b. $e^{-x}$
```{r, echo=TRUE}
x<-seq(0,3,length.out = 10)
plot(log(x), ylab = "ln(x)")

```
*20. Make a barplot of the median number of frogs grouped by whether they have spots or not.*
```{r, echo=TRUE}
plot<-barplot(c(median(dat[dat$spots==TRUE,]$frogs), median(dat[dat$spots== FALSE,]$frogs)))
axis(1, at=c(0.7,1.9), labels=c("Spots", "No Spots"))

```
*21. Plot a histogram of blue frogs*

```{r, echo=TRUE}
hist(dat[dat$color == "blue",]$frogs, xlab = "")
title("Hist of Blue Frogs")

```
*22. Use apply to calculate the across-population standard deviations in the numbers of frogs and tadpole*

```{r, echo=TRUE}
dat[,1:2] %>%
  apply(MARGIN = 2, FUN = sd)

```

*23.	Using the frog data*
  a.	Fit a linear model of tadpoles as a function of frogs for just the RED individuals and report the summary data of the fit.
  
```{r, echo=TRUE}
model<-lm(dat[dat$color == "red",]$tadpoles~dat[dat$color == "red",]$frogs)
summary(model)

```
  b.	Make a scatter plot of this data that includes the regression line
```{r, echo=TRUE}
names(model$model)<- c("tadpoles", "frogs")
r_squared<-1-sum((model$residuals)^2)/sum(model$fitted.values^2)
model$model %>%
  ggplot()+
  aes(
    x = tadpoles,
    y = frogs
  )+
  geom_point()+
  stat_summary(fun.data=mean_cl_normal) + 
  geom_smooth(method='lm')+
  annotate("text",x=3, y= 4,label=paste("r^2 = ", r_squared))
```
  c.	Fit a series of linear models of tadpoles as a function of frogs, spots, color, and their interaction terms. Build up from a simple model to the most complex model that is supported by the available data (i.e. all terms should be significant). Also test the full model that includes all variables and interaction terms.
  
```{r, echo=TRUE}
model_frogs<-lm(tadpoles~frogs,dat)
summary(model_frogs) #simple, frogs significant

model_spots<-lm(tadpoles~spots,dat)
summary(model_spots) #simple, not significant

model_color<-lm(tadpoles~spots,dat)
summary(model_color) #simple, not significant

model_frogs_spots<-lm(tadpoles~frogs:spots,dat)
summary(model_frogs_spots) # more complex, significant

model_frogs_color<-lm(tadpoles~frogs:color,dat)
summary(model_frogs_color) # more complex, significant

#Full Model
model_frogs<-lm(tadpoles~.*.,dat)
summary(model_frogs)
```
*24.	Write an if statement that makes a scatter plot of x if all the values are positive, and plots a histogram otherwise.*

```{r, echo=TRUE}
series_neg<--30:30
serise_pos<-1:60

if(all(serise_pos>0)){
  plot(serise_pos)
}else{
  hist(series_pos)
}


if(all(series_neg>0)){
  plot(series_neg)
}else{
  hist(series_neg)
}
```

*25.	Convert the more complicated graphing example at the end of  “Classic Tests” into a function that will make the same plots for any data set. Show how you would call the function passing it just the subset of data where there are 5 or more tadpoles.*

```{r, echo=TRUE}

### Plotting Example
reg1 = lm(tadpoles ~ frogs,data=dat)   #model syntax: y ~ x
plot(dat$frogs,dat$tadpoles,
     cex=1.5,  			# increase the symbol size
     col=as.character(dat$color),	# change the symbol color by name
     pch=dat$spots+1,			# change the symbol (by number)
     cex.axis=1.3,			# increase the font size on the axis 
     xlab="Frog Density",		# label the x axis
     ylab="Tadpole Density",		# label the y axis
     cex.lab=1.3,			# increase the axis label font size
     main="Frog Reproductive Effort", # title
     cex.main=2			# increase title font size
     )
abline(reg1,col="green",		# add the regression line
     ,lwd=3)				# increase the line width
legend("topleft",
       c("Red no spot","Blue no spot","Red spots","Blue Spots"),
       pch=c(1,1,2,2),
       col=c("red","blue","red","blue"),cex=1.3
       )


tess_ggplot_plot<-function(dat, x, y, color = NULL, shape = NULL, formula = NULL, title = NULL, xlab = NULL, ylab = NULL, labnames = NULL){


ggplot(data = dat) +
    aes(
      y = y,
      x = x,
      color = color, 
      shape = shape
    )+
geom_point(show.legend = TRUE)+
geom_abline(slope = formula$coefficients[2] , intercept = formula$coefficients[1])+
ggtitle(title)+
ylab(ylab)+
xlab(xlab)+
scale_shape_manual (values = c(1, 2), name = labnames)
  
}

tess_ggplot_plot(dat = dat, x = dat$frogs, y = dat$tadpoles, color = dat$color, shape = dat$spots,title = "Frog Reproductive Effort" , ylab ="Tadpole Density", xlab = "Frog Density", formula = reg1, labnames = "Spots")


#### With different data

new_data<-as.data.frame(cbind(c(1,2,3,4), c(4,3,2,1), c("black", "grey", "white", "black"), c(FALSE, TRUE, FALSE, TRUE)))

reg2<-lm(new_data$V1 ~ new_data$V2)

tess_ggplot_plot(dat = new_data , x = new_data[,1], y = new_data[,2], color = new_data[,3], shape = new_data[,4], formula = reg2, title = "Lorum Ipsum", ylab = "dolor sit ", xlab = "amet", labnames = "consectetur adipiscing elit"  )


```

*26.	Starting with a vector x = 1:10, write a for loop that adds 5 to each value in the vector. Note that it is permissible to assign an element in a vector to itself (e.g. x[i] = x[i] + 5)*
```{r, echo=TRUE}
x<-1:10

for(i in 1:10){
  x[i] <- x[i] + 5
}

```

*27.	Modify the logistic growth code to simulate discrete-time exponential growth and plot the output.*

Based off of formula $A = Pe^{rt}$:
```{r, echo=TRUE}

NT = 1000  		## number of time steps
P = 1				## initial population size
r = 0.2				## population growth rate
A = rep(0,NT)
for(i in 2:NT){
  P<-P+A[i-1]
	A[i] =  P*exp(r*i)  ## discrete exponential growth
}
plot(A)
plot(A[1:4]) #Because it's really hard to see the exponential growth over long time periods

```