---
title: "Tess McCabe's Lab 1"
output: html_notebook
---

Promt 1:
```{r, echo=TRUE}
3+12  		    ## addition
5 - pi			  ## subtraction
2*8			      ## multiplication
14/5			    ## division
14 %/% 5		  ## integer division
14 %% 5		    ## modulus (a.k.a. remainder)
sqrt(25)		  ## square root
z = 5			    ## assignment to a variable
z			        ## value of a variable
y = sqrt(36)	## square root
y
```

Prompt 2: 

```{r, echo=TRUE}
10^2  			        ## power
exp(1)				      ## exponential
log(10)			        ## natural logarithm (i.e. ln)
log10(10)			      ## log base 10
log2(10)            ## log base 2
sin(pi/2)			      ## sine
cos(pi)			        ## cosine
tan(pi/4)			      ## tangent
asin(0.5)			      ## arc-sine
acos(0.5)			      ## arc-cosine
atan(1)*180/pi			## arc-tangent
atan2(-1,-1)*180/pi	## arc-tangent (alternate version)
factorial(10)			  ## factorial
```

Prompt 3: 

1.	Evaluate the following:
   a.	ln(1)
```{r, echo=TRUE}
log(1)
```
   b.	ln(0)
```{r, echo=TRUE}
log(0)
```
   c.	ln(e)
```{r, echo=TRUE}
log(exp(1)) 
```
   d.	ln(-5)
```{r, echo=TRUE}
log(5*(-1)) # Can't do that. 
```
   e.	-ln(5)
```{r, echo=TRUE}
-log(5)
```
   f.	ln(1/5)
```{r, echo=TRUE}
log(1/5)
```



g.	How does R represent when the output of a function is not a number?

R stores things in objects. If the output of a function isn't a number, it could be text output ie: 

```{r, echo=TRUE}
print("Hello world")
```

The output could be an object that stores numbers in an ordered way like a vector, matrix, dataframe or array ie: 
```{r, echo= TRUE}
vector<-colSums(iris[,1:3], na.rm = TRUE,dims = 1)
vector
```

The output could also be a comniation of any of those things in the form of a list, or a custom object class. 


2.	Why are we multiplying the results of the atan and atan2 functions by 180/pi?

Becuase R default functions take radians ot degrees, and we wanted our answers in degrees. 

3.	How is the atan2 function different from the atan function? 

The atan2() function retunrs the angle between the x axis an the line that conencts the origin to the  (x,y) coordinates. It's the same as atan(y/x). 

4.	What is the difference between log and log10?

log() the R funnction is what most people think of as natural log  aka $ln()$. log10 is log base ten, or what most people think of when they see $log(x)$. 

5.	Given a right triangle with sides x = 5 and y = 13, calculate the length of the hypotenuse (show code)
$$x^2+y^2=c^2$$

```{r, echo=TRUE}
x<-5
y<-13
c_squared<-5^2+13^2
c<-c_squared^(1/2)
c
```

6.	If a population starts with a density of 5 individuals/hectare and grows exponentially at a growth rate r=0.04/year, what is the size of the population in π years? (show code)

```{r, echo=TRUE}
density<-5 #indiv/hect
rate<- 0.04 # year^-1
years<-pi

population<-density+(rate*years)
population
```

7.	Subtract the month you were born from the remainder left after you divide the year you were born by the day you were born (show code)

```{r, echo=TRUE}
day_born<-23 #Fyi, tomorrow is my birthday. 
month_born<-1
year_born<-1995 
remainder<-1995%%23
remainder
month_born-remainder
```