---
title: 'Homework 1'
author: "Sawyer Balint"
date: "01/31/2025"
output:
  bookdown::pdf_document2: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(bookdown) #personal style preference
library(tidyverse) #for everything
library(knitr) #for kable
```


First, let's observe the data:

```{r}
counts <- c(5,6,7,3,6,5,8,4,4,3)

hist(counts)

```

**What is your process model?**

The process model is $f(x)=\lambda$

**What is your data model?**

Because this is count data, it must be discrete and have a lower bound of 0. It also appears to have a higher density at lower values. Thus, it is likely a Poisson distribution: $$x~Pois(\lambda)$$

**Solve for the analytical MLE**
\begin{center}
Likelihood is equal to the Poisson distribution:
$$L=\Pi~Pois(x_{i}|\lambda)$$
which has the probability distribution function of:
$$Pois(x_{i}|\lambda) = \lambda^{x_i}~e^{-\lambda}$$
thus likelihood is equal to:
$$L\propto\Pi~\lambda^{x_i}~e^{-\lambda}$$
We solve for $L$ by taking the natural log:
$$lnL\propto\sum~x_iln(\lambda)-\sum(\lambda)$$
Take the derivative with respect to $\lambda$:
$$\frac{dlnL}{d\lambda}=\frac{1}{\lambda}\sum~x_i-n$$
set it equal to zero:
$$0=\frac{1}{\lambda}\sum~x_i-n$$
and solve for $\lambda$
$$\lambda=\frac{1}{n}\sum~x_i$$
Thus, $\lambda$ is equal to the sum of $x$ over $n$ which is the mean.
\end{center}

**What is the estimate for this population?**

```{r}

round(mean(counts))

```